---
title: "STAT581A4_HW2"
author: "Sam Armstrong"
date: "9/26/2019"
output: pdf_document
---
Load Libraries
```{r}
library(car)
```

1. Consider the data matrix

```{r}
X = matrix(c(-1, 3, -2, 2, 4, 2, 5, 2, 3), nrow=3, ncol=3, byrow = TRUE);X
```

(a) Calculate the matrix of deviations (residuals),X-1xbar.

```{r}
ones = rep(1, 3);
xbar = 1/3 * t(X) %*% ones;
residuals = X - t(xbar %*% ones); residuals 
```
(b) Determine S and calculate the generalized variance |S|
```{r}
S <- cov(X); S
gen.S = det(S); gen.S
```

(c) Using the results in (b), calculate the total sample variance.

```{r}
totalSampleVar = sum(diag(cov(X))); totalSampleVar
```

2*. In  northern  climates,  roads  must  be  cleared  of  snow  quickly  following  a  storm.   One  measure  of storm severity is x1 = its duration in hours, while the effectiveness of snow removal can be quantified by x2 = the number of hours crews, men, and machine, spend to clear the snow.  Here are the results for 25 incidents in Wisconsin. Calculate the mean vector, the covariance matrix, and the correlation matrix. Also, make a scatterplot and describe the relationship between the variables.

```{r}
snow = matrix(c(12.5, 13.7, 14.5, 16.5, 8.0, 17.4, 9.0, 11.0, 19.5, 23.6, 8.0, 13.2, 9.0, 32.1, 7.0, 12.3, 7.0, 11.8, 9.0, 24.4, 6.5, 18.2, 10.5, 22.0, 10.0, 32.5, 4.5, 18.7, 7.0, 15.8, 8.5, 15.6, 6.5, 12.0, 8.0, 12.8, 3.5, 26.1, 8.0, 14.5, 17.5, 42.3, 10.5, 17.5, 12.0, 21.8, 6.0, 10.4, 13.0, 25.6), nrow=25, ncol=2, byrow = TRUE)
colnames(snow) <- c('x1', 'x2')
snow
```
```{r}
ones = rep(1, 25);
xbar = colMeans(snow); xbar
covMatrix = cov(snow); covMatrix
corMatrix = cor(snow); corMatrix
plot(snow)
```
The two vairiables are positivly correlated by a magnitude of 0.45 suggesting that increasing the value of one variable increases the value of the other.

3. Evaluate T2 for testing H0:mu=[7 11], using the data:

```{r}
mu0 = c(7, 11)
X = matrix(c(2, 12, 8, 9, 6, 9, 8, 10), nrow=4, ncol=2, byrow = TRUE);X
sampleMean = apply(X, 2, mean);
S  = cov(X);
TSqd = 4*t(sampleMean-mu0)%*%solve(S)%*%(sampleMean-mu0);TSqd
```

4*. A physical anthropologist performed a mineral analysis of nine ancient Peruvian hairs.  The results for chromium (x1) and strontium (x2) levels, in parts per million (ppm), were as follows: 

```{r}
hair = matrix(c(0.48, 40.53, 2.19, 0.55, 0.74, 0.66, 0.93, 0.37, 0.22, 12.57, 73.68, 11.13, 20.03, 20.29, 0.78, 4.64, 0.43, 1.08), nrow=2, ncol=9, byrow = TRUE)
rownames(hair) <- c('x1 (Cr)', 'x2 (St)')
hair = t(hair); hair
```
It is known that low levels (less than or equal to 0.100ppm) of chromium suggest the presence of diabetes, while strontium is an indication of animal protein intake.

(a) Construct and plot a 90% joint confidence ellipse for the population mean vector for mu=[mu1,mu2]T,assuming that these nine Peruvian hairs represent a random sample from the individuals belonging to aparticular ancient Peruvian culture.

```{r}
alpha = 0.9
p = 2
n = nrow(hair);
center = apply(hair, 2, mean)
S = cov(hair)
r.val = sqrt((n -  1) * p/(n * (n - p)) * qf(alpha, p, n - p));
plot(ellipse(center = center, shape = S, radius = r.val, draw = FALSE), type="l", xlab = 'x1 (CR)', ylab = 'x2 (ST)', xlim = range(hair[,1]), ylim = range(hair[,2]))
points(hair)
points(center[1], center[2], col = 'red', cex = 2, pch = 4)
```

(b) Test the hypothesis that mu=[0.30 10]T. Discuss whether this mean vector is a plausible value for mu.

```{r}
mu0 = c(0.30, 10)
sampleMean = apply(hair, 2, mean)
S  = cov(hair);
TSqd = 9*t(sampleMean-mu0)%*%solve(S)%*%(sampleMean-mu0);TSqd
critVal0.10 = 8*2/9*qf(0.90, 2, 9);critVal0.10
```
Since the Tsqd value (1.772481) is less than the critical value (5.344804) we fail to reject that the vector [0.30, 10]T is a plausible value of mu. You can check this by seeing if the point (0.30, 10) is in the confidense ellipse which it is. 


(c) Do these data appear to be bivariate normal?  Discuss their status with reference to Q-Q plots anda scatter diagram. If the data are not bivariate normal, what implications does this have for the results of (a) and (b)?

```{r}
qqnorm(hair);qqline(hair, col = 2)
plot(hair)
shapiro.test(hair)
```
No the data does not appear to be bivariate normal due to the long tails and poor fit in the Q-Q Plots and the non-linear looking spread of data in the scatter diagram. This is also confirmed from running the Shapiro-Wilk Normality test and getting a p-value of 9.484e-06 which rejects the null hypothesis. Since the data is not bivariate normal the results from (a) and (b) are possibly innaccurate/not valid.

(d) Repeat the analysis with the obvious “outlying” observation removed. Do the inferences change? Comment.

```{r}
hair2 = matrix(c(0.48, 2.19, 0.55, 0.74, 0.66, 0.93, 0.37, 0.22, 12.57, 11.13, 20.03, 20.29, 0.78, 4.64, 0.43, 1.08), nrow=2, ncol=8, byrow = TRUE)
rownames(hair2) <- c('x1 (Cr)', 'x2 (St)')
hair2 = t(hair2); hair2
alpha = 0.9
p = 2
n = nrow(hair2);
center = apply(hair2, 2, mean)
S = cov(hair2)
r.val = sqrt((n-  1) * p/(n * (n - p)) * qf(alpha, p, n - p));
plot(ellipse(center = center, shape = S, radius = r.val, draw = FALSE), type="l", xlab = 'x1 (CR)', ylab = 'x2 (ST)', xlim = range(hair2[,1]), ylim = range(hair2[,2]))
points(hair2)
points(center[1], center[2], col = 'red', cex = 2, pch = 4)
mu0 = c(0.30, 10)
sampleMean = apply(hair2, 2, mean);
S  = cov(hair2);
TSqd = 8*t(sampleMean-mu0)%*%solve(S)%*%(sampleMean-mu0);TSqd
critVal0.10 = 7*2/8*qf(0.90, 2, 8);critVal0.10
qqnorm(hair2);qqline(hair2, col = 2)
plot(hair2)
shapiro.test(hair2)
```
Since the Tsqd value (5.307872) is less than the critical value (5.447956) we fail to reject that the vector [0.30, 10]T is a plausible value of mu. You can check this by seeing if the point (0.30, 10) is in the confidense ellipse which it is. 

No the data does not appear to be bivariate normal due to the long tails and poor fit in the Q-Q Plots and the non-linear looking spread of data in the scatter diagram. This is also confirmed from running the Shapiro-Wilk Normality test and getting a p-value of 8.413e-05 which rejects the null hypothesis. Since the data is not bivariate normal the results from (a) and (b) are possibly innaccurate/not valid.

After removing the outlier the inferences do not change. This makes sense since removing one data point shouldn't cause the whole data set (size larger than 8) to become bivariate normal.


5*. Municipal waster water treatment plants are required by law to monitor their discharges into rivers and streams on a regular basis.  Concerns about the reliability of data from one of these self-monitoring programs led to a study in which samples of effluent were divided and sent to two laboratories for testing. One-half of each sample was sent to the Wisconsin State Laboratory of Hygiene, and one-half were sent to a private commercial laboratory routinely used in the monitoring program.  Measurements of biochemical oxygen demand (BOD) and suspended solids (SS) were obtained for n= 11 sample splits, from the two laboratories.

```{r}
clab = matrix(c(6, 27, 6, 23, 18, 64, 8, 44, 11, 30, 34, 75, 28, 26, 71, 124, 43, 54, 33, 30, 20, 14), nrow=11, ncol=2, byrow = TRUE)
colnames(clab) <- c('x1j1 (BOD)', 'x1j2 (SS)')
print('Commercial lab')
clab

slab = matrix(c(25, 15, 28, 13, 36, 22, 35, 29, 15, 31, 44, 64, 42, 30, 54, 64, 34, 56, 29, 20, 39, 21), nrow=11, ncol=2, byrow = TRUE)
colnames(slab) <- c('x2j1 (BOD)', 'x2j2 (SS)')
print('State lab of hygiene')
slab
```

Do the two laboratories’ chemical analyses agree? If differences exist, what is their nature? Hint:  First begin by calculating the differences in the measurements between the two labs.  For instance,d11=x111-x211= 6-25 =-19.  Calculate a T2-statistic and test H0:d=[d1 d2]T=[0 0]T.

```{r}
xbar1 = colMeans(clab);xbar1
xbar2 = colMeans(slab);xbar2
S1 = cov(clab);S1
S2 = cov(slab);S2
dataDifference = clab-slab;dataDifference
difference = xbar1-xbar2; difference
S = ((10*S1) + (10*S2))/20;S
Tsqd = t(difference) %*% solve((2/11) * S) %*% difference;Tsqd
critVal = (22-2)*2/(22-2-1)*qf(.95, 2 , 19);critVal
```
No the two labs chemical analysis do not agree. The average difference is too high ( -9.363636  13.272727) and the difference between the datasets is high and all over the place (both positive and negative differences). Reject the null hypotheis since the Tsqd value (12.6648) is greater than the critical value (7.414512) at alpha=0.05.