---
title: "STAT581A4_HW3"
author: "Sam Armstrong"
date: "10/15/2019"
output: pdf_document
---
```{r}
library(car)
skull <- read.csv(file="C:/Users/Sam/STAT581_A4/skull.csv", quote = "'")
```
1)
```{r}
manovaOut = manova(as.matrix(skull[,1:4])~as.factor(skull[, "Era"]), subset = as.factor(skull[, "Era"]) %in% c(1, 2));
summary(manovaOut)
```
FTR the null hypothesis: mean_era1 = mean_era2. Yes it does seem reasonable to pool variances on skulls from the two different eras.

2.a)
```{r}
x1=c(6, 5, 8, 4, 7, 3, 1, 2, 2, 5, 3, 2)
x2=c(7, 9, 6, 9, 9, 3, 6, 3, 3, 1, 1, 3)
x3=c(1, 1, 1, 1, 1, 2, 2, 2, 3, 3, 3, 3)
x = data.frame(x1, x2, x3)

m = manova(as.matrix(x[,1:2])~factor(x[,3]))
summary(m, test="Wilks")
```

2.b) 
```{r}
S1 = var(x[x[,3] == 1, 1:2]); 
S2 = var(x[x[,3] == 2, 1:2]);
S3 = var(x[x[,3] == 3, 1:2]);
n1 = sum(x[,3] == 1);
n2 = sum(x[,3] == 2);
n3 = sum(x[,3] == 3);
W = (n1 - 1)*S1 + (n2 - 1)*S2 + (n3 - 1)*S3; 
g = 3;
p = 2;
n = 12;
T = (n-1)*var(x[, 1:2]);
B = T - W; 
A = B %*% solve(W);
wilksLambda = det(W)/det(B+W);
wilksLambda
```

```{r}
pairwiseOut1 = manova(as.matrix(x[, 1:2])~as.factor(x[, 3]), subset = as.factor(x[, 3]) %in% c(1, 2));
summary(pairwiseOut1)

pairwiseOut2 = manova(as.matrix(x[, 1:2])~as.factor(x[, 3]), subset = as.factor(x[, 3]) %in% c(1, 3));
summary(pairwiseOut2)

pairwiseOut3 = manova(as.matrix(x[, 1:2])~as.factor(x[, 3]), subset = as.factor(x[, 3]) %in% c(2, 3));
summary(pairwiseOut3)
```

No the mean vectors are not equal across the three treatments (the p-value when comparing the mean vectors of treatment 1 with treatment 2 (0.0009507) and treatment 1 with treatment 3 (5.549e-05) were both less than 0.05).

2.c)
```{r}
t.stat = -1*(n-1-(p+g)/2)*log(wilksLambda); t.stat
crit.val = qchisq(.95, p*(g-1)); crit.val
pval = 1 - pchisq(t.stat, p*(g-1));pval
```
Since the t.stat (28.21136) > crit.val (9.487729) we reject the null hypothesis. There is evidense at alpha=0.05 (> p-value 1.13011e-05) that the mean treatments characteristics are not all equal across treatments.